---
title: Analysis of admission in elementary school in 2014 in Romania
output:
  html_document:
    number_sections: false
    toc: true
    fig_width: 8
    fig_height: 6
    theme: cosmo
    highlight: tango
    code_folding: show
---

```{r setup, include=FALSE}
library(jsonlite)
library(rgdal)
library(tidyr)
library(leaflet)
require(plotrix)
library(plyr)
library(dplyr)
library(ggplot2)
library(knitr)
library(kableExtra)
options(knitr.table.format = "html") 
```


# **Introduction**


In a previously developed [Kernel](https://www.kaggle.com/gpreda/esar-2014-exploratory-analysis), we explored the data in the pupils file, focusing on migration from rural to urban areas, type of education, how the disabled pupils have access in rural and urban area to other educational options than the main-stream, traditional education system and also relationship between ethnic apartenence, disabled status and orphan or institutionalized status for the pupils registered in our data file.  
In this Kernel we will use the data in pupils, schools and 2 geolocation files to represent on a *leaflet*  map the percent of population (values from last census of 2011) that registered as pupils in 1st class in elementary school in 2014. To calculate this percent, we will merge all  4 sources of data, using as join factors the SIRUES code (for join between pupils data and schools data) and the county name (for join between the schools+pupils data  and geojeson data and between the two geojson data). As the county names are written  differently in the 3 data files, we will use a function for matching the county names with fuzzy match based on Levenshtein distance between the compared strings. 


# **Function matchingNames**


The function 'matchingNames' is used to do fuzzy matching between Romanian counties names from data files where diferent encoding or writing of these names makes simple matching not  effective.

```{r matching_names}
# Matching names using: https://www.r-bloggers.com/fuzzy-string-matching-a-survival-skill-to-tackle-unstructured-information/
#match using Levenshtein distance between the compared strings 
matchingNames <- function(source1, source2) {
  # To make sure we are dealing with characters
  source1<-as.character(source1)
  source2<-as.character(source2)

  # It creates a matrix with the Standard Levenshtein distance between the name fields of both sources
  dist.name<-adist(source1,source2, partial = FALSE, ignore.case = FALSE)
  # We now take the pairs with the minimum distance
  min.name<-apply(dist.name, 1, min)

  match.s1.s2<-NULL  
  for(i in 1:nrow(dist.name))
  {
    s2.i<-match(min.name[i],dist.name[i,])
    s1.i<-i
    match.s1.s2<-rbind(data.frame(s2.i=s2.i,s1.i=s1.i,
                                  s2name=source2[s2.i], s1name=source1[s1.i], 
                                  adist=min.name[i]),match.s1.s2)
  }

  
  mapi<-c(1:nrow(dist.name))
  for(j in 1:nrow(dist.name))
  {
    mapi[match.s1.s2$s1.i[j]] = match.s1.s2$s2.i[j]
  }
  return(mapi)
}
```


# **Read pupils data**



From the pupils data we extract only a part of the columns, including sex, pupils environment (rural or urban), ethnicity (in Romania there are multiple ethnic groups, with most notable besides Romanian the Hungarian minorities, the Rroma (Gipsies), German, Ucrainian ...), educational alternative (beside traditional, special education, Waldorf, Step-by-step ...), teaching language, SIRUES code,  school environment (either rural or urban), handicap (if the pupils registered has a certain permanent health condition), orphan or institutionalized pupils flag, single parent flag.  
SIRUES codes are  used to uniquely identify the schools.


```{r read_pupils_info}
#PATH="../input/elementary-school-admission-romania-2014/"
PATH="../input/"

#read data about pupils registering in 1st grade in 2014
#the data contains SIRUES code (unique for each school) - we will use this code to merge pupil's data with schools data
raw.data <- read.csv(paste(PATH,"elementary_school_registration_2014.csv",sep=""))
#perform some data cleansing, reducing some irrelevant columns and renaming columns for clarity
# 'SIRUES' is an unique code attributed to each school
pupils<-cbind(raw.data[,c(2,3,5,7,8,9,10,11,12,13,14)])
column.names.pupils<-c("sex","origin_env","ethnic","admission_stage","educational_alternative","teaching_language","SIRUES",
                       "school_env","disabled","orphan","single_parent")
colnames(pupils)<-column.names.pupils
```


# **Read schools data**


From the schools data we keep only few dimmensions: the county (`judet`) name, the SIRUES code  (this code will be used to match the information from pupils data file), the school type  and the school category.


```{r read_schools_data}
#read data about schools
#'judet' is an administrative unit in Romania, like 'county' in US
raw.SIRUES <- read.csv(paste(PATH,"school_network.csv",sep=""))
#perform some data cleansing, reducing some irrelevant columns and renaming columns for clarity
schools<-cbind(raw.SIRUES[,c(2,3,4,6,7)])
column.names.schools<-c("judet","name","SIRUES","school_type","school_category")
colnames(schools)<-column.names.schools
```

# **Merge pupils and school information**


Merging the pupils information with school information is done using the SIRUES code.


```{r merge_pupils_schools}
#merge schools and pupils data on SIRUES common code
pupils.schools <- merge(pupils, schools, by="SIRUES")
```


After merging, the data with pupils and school information is kept in `pupils.schools` data frame.


## Teaching language



Let's see what is the teaching language for each ethnic group.


```{r ethnic_language}
pupils %>% 
  group_by(ethnic, teaching_language) %>% summarise(nums=length(SIRUES)) %>% ungroup() %>%
    complete(ethnic, teaching_language, fill = list(nums = 0)) -> ethnic_language

ethnic_language %>%
  group_by(ethnic) %>%
  mutate(all = sum(nums),  perc = round(100 * nums/all, 2)) -> ethnic_language

ggplot(ethnic_language, aes(teaching_language, ethnic)) + geom_tile(aes(fill = perc),
      colour = "white") + scale_fill_gradient(low = "lightblue", high = "blue",name = "Percent\nusage") + 
      theme_grey() + labs(x = "", y = "") + 
      labs(title="Language used by each ethnic group", subtitle="Percent of usage for language is shown per ethnic group") +
      scale_x_discrete(expand = c(0, 0)) +  scale_y_discrete(expand = c(0, 0)) +
      theme(axis.text.x = element_text(angle = 90, hjust = 1), axis.ticks = NULL)
```


## Pupil environment (rural vs. urban) per county


Let's see what is the percent of rural/urban 1st grade pupils (origin) population per each county.


```{r fig.width=12, fig.height=5,pupil_env_rural_urban_population_per_county}
pupils.schools %>% 

  group_by(judet, origin_env) %>% summarise(nums=length(SIRUES)) %>% ungroup() %>%
    complete(judet, origin_env, fill = list(nums = 0)) -> county_environment

county_environment %>%
  group_by(judet) %>%
  mutate(all = sum(nums),  perc = round(100 * nums/all, 2)) -> county_env

ggplot(county_env, aes(judet, origin_env)) + geom_tile(aes(fill = perc),
      colour = "white") + scale_fill_gradient(low = "lightgreen", high = "darkblue",name = "Percent") + 
      theme_grey() + labs(x = "County", y = "Pupil environment: Rural(R) Urban (U)") +  
      labs(title="Pupil environment for each county", subtitle="Percent of pupil environment for each county (rural/urban)") +
      scale_x_discrete(expand = c(0, 0)) +  scale_y_discrete(expand = c(0, 0)) +
      theme(axis.text.x = element_text(angle = 90, hjust = 1), axis.ticks = NULL)
```



## School environment (rural vs. urban) per county


Let's see what is the percent of rural/urban 1st grade places in schools per each county.


```{r fig.width=12, fig.height=5,school_env_rural_urban_population_per_county}
pupils.schools %>% 

  group_by(judet, school_env) %>% summarise(nums=length(SIRUES)) %>% ungroup() %>%
    complete(judet, school_env, fill = list(nums = 0)) -> county_sch_environment
    
county_sch_environment %>%
  group_by(judet) %>%
  mutate(all = sum(nums),  perc = round(100 * nums/all, 2)) -> county_sch_env

ggplot(county_sch_env, aes(judet, school_env)) + geom_tile(aes(fill = perc),
      colour = "white") + scale_fill_gradient(low = "lightgreen", high = "darkblue",name = "Percent") + 
      theme_grey() + labs(x = "County", y = "School environment: Rural(R) Urban (U)") + 
      labs(title="School environment for each county", subtitle="Percent of school environment for each county (rural/urban)") +
      scale_x_discrete(expand = c(0, 0)) +  scale_y_discrete(expand = c(0, 0)) +
      theme(axis.text.x = element_text(angle = 90, hjust = 1), axis.ticks = NULL)
```


## Ethnic distribution per county


Let's see what is the ethnic distribution per county.


```{r fig.width=12, fig.height=8,ethnic_distribution_per_county}
pupils.schools %>% 
  group_by(judet,ethnic) %>% summarise(nums=length(SIRUES)) %>% ungroup() %>%
    complete(judet, ethnic,  fill = list(nums = 0)) -> ethnic_county

ethnic_county %>%
  group_by(judet) %>%
  mutate(all = sum(nums),  perc = round(100 * nums/all, 2)) -> ethnic_p_county


ggplot(ethnic_p_county, aes(judet,ethnic)) + geom_tile(aes(fill = perc),
      colour = "white") + scale_fill_gradient(low = "lightgreen", high = "darkblue",name = "Percent") + 
      theme_grey() + labs(x = "County", y = "Ethnic group") + 
      labs(title="Ethnic pupils population per each county", subtitle="Percent of ethnic pupils population per county") +
      scale_x_discrete(expand = c(0, 0)) +  scale_y_discrete(expand = c(0, 0)) +
      theme(axis.text.x = element_text(angle = 90, hjust = 1), axis.ticks = NULL)
```


## School category per county


Let's see what is the school category distribution per county.


```{r fig.width=12, fig.height=8,school_category_distribution_per_county}

pupils.schools %>% 
  group_by(judet,school_category) %>% summarise(nums=length(SIRUES)) %>% ungroup() %>%
    complete(judet, school_category,  fill = list(nums = 0)) -> school_category_county

school_category_county %>%
  group_by(judet) %>%
  mutate(all = sum(nums),  perc = round(100 * nums/all, 2)) -> school_p_category_county

ggplot(school_p_category_county, aes(judet,school_category)) + geom_tile(aes(fill = perc),
      colour = "white") + scale_fill_gradient(low = "yellow", high = "darkgreen",name = "Percent") + 
      theme_grey() + labs(x = "County", y = "School category") + 
      labs(title="School category per each county", subtitle="Percent of school category (by number of registeres pupils) per county") +
      scale_x_discrete(expand = c(0, 0)) +  scale_y_discrete(expand = c(0, 0)) +
      theme(axis.text.x = element_text(angle = 90, hjust = 1), axis.ticks = NULL)
```  



Most of the schools, in all counties are `Gimnasium`, i.e. a school with both elementary school and junior high role.
In counties with relative large urban population, there is a considerable percent of `Liceu` i.e. highschools, some of them including junior-high as well. In `Hunedoara` county, there are almost equal number of `Colegiu` (College) and `Gimnasium`. 


## Educational alternative for disabled pupils  


We would like to see how different are the educational alternatives for disabled pupils from Rural and Urban areas.



### Education alternative vs. origin environment  


Let’s start with the plot of education alternative vs. origin environment (rural or urban). 


```{r fig.width=6, fig.height=4,education_alternative_vs_origin_env}
pupils %>% filter(disabled == "DA") %>% ungroup() -> pupilsDisabled

qplot(pupilsDisabled$origin_env,fill=pupilsDisabled$educational_alternative) +
theme_bw() + 
labs(title="Education alternative for disabled pupils vs. origin environment", 
     x="Origin of disabled pupils (R=Rural, U=Urban)", y="Pupils number") + 
scale_fill_discrete(name = "Educational\nalternative")
```



Disabled pupils from Rural areas are less likely to access an alternative education, compared with pupils from Urban areas. If they will follow an alternative education path, they will most probably join the Special education system, which provide some support for them but still applies some traditional education approach (**Invatamant special** stands for **Special education**). 
For children in Urban areas, the percent of them following alternative education (not the **Traditional**) is larger and also they will have access to diferent educational systems, like **Currative education**, **Step by step**, **Waldorf**, **Freinet**.




### Education alternative vs. school environment  

The following graph is showing the school environment for the disabled pupils (Rural or Urban).


```{r fig.width=6, fig.height=4,education_alternative_vs_school_env}
qplot(pupilsDisabled$school_env,fill=pupilsDisabled$educational_alternative) +  theme_bw() + 
labs(title="Education alternative for disabled pupils vs. school environment", x="School of disabled pupils (R=Rural, U=Urban)", y="Pupils number") + 
scale_fill_discrete(name = "Educational\nalternative")
```  


We can see that in the Rural areas there are zero non-traditional educational alternatives. This will also mean that the disabled pupils from country side will have to travel to the cities in order to get access to non-traditional educational alternatives.


### Focusing only on non-traditional education alternative  


We focus only on non-traditional education alternative, to see how the disabled pupils have access to such educational alternatives in countryside and city areas.


#### Education alternative vs. origin environment  



Let’s look to the education alternative used by pupils with handicap from Rural and Urban areas.


```{r fig.width=6, fig.height=4,education_alternative_nt_vs_origin_env}
pupils %>% filter(disabled == "DA", educational_alternative != "Tradi?ional?") %>% ungroup() -> pupilsDisabledNT
qplot(pupilsDisabledNT$origin_env,fill=pupilsDisabledNT$educational_alternative) +  theme_bw() + 
labs(title="Education alternative for disabled pupils vs. origin environment",
 subtitle="Non traditional options only", x="Origin of disabled pupils (R=Rural, U=Urban)", y="Pupils number") + 
scale_fill_discrete(name = "Educational\nalternative")
```  

If we remove the **Traditional** education alternative, we can see easier that pupils from country side attend only **Step by step** (few of them) and **Special education** options.


#### Education alternative vs. school environment  


Let’s look now to the education alternative available in the Rural and Urban areas.


```{r fig.width=6, fig.height=4,education_alternative_nt_vs_school_env}
pupils %>% filter(disabled == "DA", educational_alternative != "Tradi?ional?") %>% ungroup() -> pupilsDisabledNT
qplot(pupilsDisabledNT$school_env,fill=pupilsDisabledNT$educational_alternative) +  theme_bw() + 
labs(title="Education alternative for disabled pupils vs. school environment",
  subtitle="Non traditional options only", x="School of disabled pupils (R=Rural, U=Urban)", y="Pupils number") + 
scale_fill_discrete(name = "Educational\nalternative")
```  


Looking now to the schools environment (rural or urban), we see that in country side only Step by step option is available, and for a very small number of pupils, apart of the Traditional option. 
The majority of disabled pupils from country side will have to go to an urban area to attend first class in a non-traditional school.



## Disabled kids that are orphaned or institutionalized  



We will focus now on the disabled kids that are orphaned or institutionalized.



### Rural vs. urban  


First, we would like to look side by side for rural and urban environment.


```{r fig.width=6, fig.height=4,disabled_kids_institutionalized}
pupils %>% filter(disabled == "DA") %>% ungroup() -> pupilsDisabledOrphaned
qplot(pupilsDisabledOrphaned$origin_env,fill=pupilsDisabledOrphaned$orphan) +  theme_bw() + 
labs(title="Disabled pupils - comparison of origin and family status", x="Origin of disabled pupils (R=Rural, U=Urban)", y="Pupils number") + 
scale_fill_discrete(name  ="Is orphan", breaks=c("", "DA"), labels=c("NO", "YES"))
```  


One thing that we notice is that for the disabled pupils, a significant percent are as well orphan or institutionalized. 
Looking also to environment, we see that while there is significantly less disabled pupils in rural areas than in urban areas (compared with the whole population status), there are also more rural pupils (as total and percent) than urban pupils.


### Nationality of disabled pupils orpahned or instututionalized   



```{r fig.width=6, fig.height=4,disabled_kids_institutionalized_nationality}
pupils %>% filter(disabled == "DA") %>% ungroup() -> pupilsDisabledOrphaned
qplot(pupilsDisabledOrphaned$ethnic,fill=pupilsDisabledOrphaned$orphan) +  theme_bw() + 
labs(title="Disabled pupils - comparison of ethnic groups and family status", x="Ethnic group of disabled pupils", y="Pupils number") + 
scale_fill_discrete(name  ="Is orphan", breaks=c("", "DA"), labels=c("NO", "YES"))
```  



Comparing now the disabled pupils based on nationality, we also see an unbalance, compared with the whole population status.  
The **Rroma** ethnic group shows a very large percent of institutionalized or orphan disabled pupils, while percents of institutionalized or orphan disabled pupils is significantly large as well in **Hungarian** and still very large in **Romanian* ethnic  group. There are no other ethnic groups with registered disabled orhpans or institutionalized pupils.  


### Gender of disabled pupils orpahned or instututionalized  


```{r fig.width=6, fig.height=4,disabled_kids_institutionalized_gender}
pupils %>% filter(disabled == "DA") %>% ungroup() -> pupilsDisabledOrphaned
qplot(pupilsDisabledOrphaned$sex,fill=pupilsDisabledOrphaned$orphan) +  theme_bw() + 
labs(title="Disabled pupils - comparison of gender and family status", x="Gender of disabled pupils", y="Pupils number") + 
scale_fill_discrete(name  ="Is orphan", breaks=c("", "DA"), labels=c("NO", "YES"))
```

We also observe a large inbalance of genders of disabled pupils. There are much more disabled boys than girls. 
As for institutionalized or orphans, we can see that as a percent, disabled girls are more frequently orphan or institutionalized (although in absolute numbers are less).  


# **Read geojson information**

There are two geojson data files; the first is used for rendering; the second is used for extracting the census information. While the first has a reduced number of points and in the same tme the counties boundaries are represented with a good-enough accuracy for our purpose, the second file have a too large number of contour points and will be used just to extract the census data.  We read both geojson files.

```{r geojson_data}
#geoJSON used for rendering, has lower number of points
rgeojson <- readOGR(paste(PATH,"romania.geojson",sep=""), layer="OGRGeoJSON")
#geoJSON used for retrieving population information from census
geojson <- readLines(paste(PATH,"ro_judete_poligon.geojson",sep=""), warn = FALSE, encoding = "UTF-8") %>%
  paste(collapse = "\n") %>%
  fromJSON(simplifyVector = FALSE)
```

The geojson for rendering is stored in `rgeojson` and the geojson with the census data is stored in `geojson`.


## List with census information


We create a list with census information for all census years, with information for every county ('judet'). We then extract the population for last census (2011). This will be used to calculate the percent from each county population that is registered in 2014 for 1st year in elementary school.


```{r census_population}
# Gather population for all counties ('judet')
poplist <- lapply(geojson$features, function(feat) {
  c(as.character(feat$properties$name),as.integer(feat$properties$pop1948),as.integer(feat$properties$pop1956),as.integer(feat$properties$pop1977), as.integer(feat$properties$pop1992), as.integer(feat$properties$pop2002),as.integer(feat$properties$pop2011))

})
census.years<-c("1948", "1956","1977", "1992", "2001", "2011")
pop <- matrix(unlist(poplist), nrow = 42, ncol = 7, byrow = TRUE)
colnames(pop) <- c("judet",census.years)
population2011 <- pop[,c(1,7)]
```  

## Join the pupils and schools data with geojson data

We join the data from pupils and schools with the geojson data using a fuzzy match with *matchingNames function*. Join column is the `county` (**judet**) name. As the name is not standardized (in one version diacritics are used, while in other not, some versions uses hyphen while others not) only the fuzzy match allows a fully-automated processing of the data.

```{r fuzzy_match_for_join}
#get pupils registered per each county
pupils_registered_per_county <- plyr::count(pupils.schools, "judet")
#match the counties with 'judet' (ordering as in the geojson)
pupilsCounty <- pupils_registered_per_county[matchingNames(rgeojson$name,as.character(pupils_registered_per_county$judet)),]
populationCounty <-population2011[matchingNames(rgeojson$name,as.character(population2011[,1])),]
```

## Percent of population registered for 1st grade elementary school


The percent of the population registered for 1st grade elementary school in 2014 is approximated, from pupils data and census data, as the report of number of pupils registered per county in 2014 and last population estimate during the census of 2011.  
All the pupils registering in 2014 were already born in 2011.


```{r prepare_percent}
percentPupilsCounty <- pupilsCounty$freq / as.integer(populationCounty[,2])
# Color by number of pupils admited
displayPercent = round(percentPupilsCounty * 100,2)
pal <- colorBin("Greens", displayPercent, bins = 8, na.color = "Red")
```


# **Interactive maps**


## Map with percent of population registered for 1st grade elementary school


The interactive map shows each county with:

* the number of pupils registering 1st grade in 2014
* the population in 2011
* percentage of pupils in 1st grade (2014) from the total population (2011 census)

 


Hint: <font color="green"><b>click on the map</b></font> to see detailed information on each county.  



```{r percent_of_pupils}
#initialize popup
countyPopup <- paste0("<h3>Judet (County):&nbsp<font color=\"red\">",pupilsCounty$judet,"</font></h3>",
                      "Pupils registered for 1st grade, 2014:&nbsp<b><font color=\"red\">",pupilsCounty$freq,"</font></b>",
                      "<br><br></strong>Population, 2011:&nbsp<b><font color=\"red\">",populationCounty[,2],"</font></b>",
                      "<br><br></strong>Percent pupils in 1st grade (2014)<br>from population (2011 census):&nbsp<b><font color=\"red\">",displayPercent,"%</font></b>")

#prepare leaflet
leaflet(data = rgeojson) %>% 
  addTiles() %>%
  setView(25.6, 45.6, 7) %>% 
  addPolygons(fillColor = ~pal(displayPercent), 
              fillOpacity = 0.9, 
              color = "#7700BB", 
              weight = 1, 
              popup = countyPopup) %>%
  addLegend("topright", pal = pal, 
            values = c(min(displayPercent),
                       max(displayPercent)),
            title = "Percent pupils registered<br>in 1st grade (2014)<br> from population (2011 census)",
            labFormat = labelFormat(suffix = "%"),
            opacity = 1
  )
```  


The map shows large regional variations from the counties in Moldova to the counties in the South of Romania.


## Map with number of schools per county


Hint: <font color="blue"><b>click on the map</b></font> to see detailed information on each county.  


```{r number_of_schools_per_county}

#get number of schools per each county
schools_per_county <- plyr::count(schools, "judet")
#match the counties with 'judet' (ordering as in the geojson)
schoolsCounty <- schools_per_county[matchingNames(rgeojson$name,as.character(schools_per_county$judet)),]
nSchoolsCounty = schoolsCounty$freq
pal <- colorBin("Blues", nSchoolsCounty, bins = 8, na.color = "Red")
#initialize popup
countySchoolsPopup <- paste0("<h3>Judet (County):&nbsp<font color=\"blue\">",schoolsCounty$judet,"</font></h3>",
                            "<br></strong>Number of schools:&nbsp<b><font color=\"blue\">",nSchoolsCounty,"</font></b>"
                      )

#prepare leaflet
leaflet(data = rgeojson) %>% 
  addTiles() %>%
  setView(25.6, 45.6, 7) %>% 
  addPolygons(fillColor = ~pal(nSchoolsCounty), 
              fillOpacity = 0.9, 
              color = "#77AABB", 
              weight = 1, 
              popup = countySchoolsPopup) %>%
  addLegend("topright", pal = pal, 
            values = c(min(nSchoolsCounty),
                       max(nSchoolsCounty)),
            title = "Number of schools<br>per county",
            labFormat = labelFormat(suffix = ""),
            opacity = 1
  )
```  


# **Conclusions**

We represented the percent of the population (from county population) registered as pupils in 1st grade of  elementary school in 2014. The data shows a very uneven distribution at county level (from county estimated population and school admission registration) of percent of pupils from total population. The values might be related with either variable number of children in families from various counties, with results of different modalities of migration (some counties population might favor letting children back with relatives while parents are emigrating to work in EU countries, another counties population might take the children when leaving the country).


# **Feedback**

Thank you for reading this Kernel. I will appreciate your **comments and suggestions*.


# **References**

[1] Public data from Romanian Government, http://data.gov.ro/   
[2] Romanian geo datasets, http://www.geo-spatial.org/download/romania-seturi-vectoriale  
[3] SIRUES codes Romania 2014, http://eprofu.ro/docs/tehnic/institutii/retea-scolara.xls   
[4] School cartography, https://www.siiir.edu.ro/carto/  
[5] Fuzzy string matching, https://www.r-bloggers.com/fuzzy-string-matching-a-survival-skill-to-tackle-unstructured-information/  
