---
title: Romanian Referendum 2018
output:
  html_document:
    number_sections: false
    toc: true
    fig_width: 8
    fig_height: 6
    theme: cosmo
    highlight: tango
    code_folding: hide
---

# **Introduction**


## About the Referendum

We present here the partial results of the Romanian Referendum 2018 (6-7 October). 

The subject of the Referendum is related with banning same-sex marriages in Romania, by Constitution.  

Presently, same-sex marriages are not permitted in Romania, nevertheless, there was a confederation of religious organizations that promoted this initiative.

In order to be validated, 30% of the total registered electors should vote. Each day the vote is open from 7 am to 9 pm.

## About Romania  

Romania is a country part of the European Union (from 2007). Its population is estimated to 21.5 millions people (2017) and current number of registered voters is 18.27 millions people.  

Romania has 41 counties (*judet* in Romanian) and the capital city, Bucharest, totally 42 administrative units.  



# **Load the data**


## Packages


```{r setup, include=FALSE}
library(plotly)
library(dplyr)
library(gtools)
library(gridExtra)
library(grid)
library(jsonlite)
library(rgdal)
library(tidyr)
library(leaflet)
library(ggrepel)
```

## Data files

The data is provided by the Romanian Electoral Bureau. They report the partial presence to vote on the two days, at 10 am, 13 pm, 16 pm, 19 pm and 21 pm.


```{r read_referendum_data,warnings=FALSE}
PATH = "../input/referendum-ro-2018/"
presence_0610 <- read.csv(paste(PATH,"presence_RO_2018-10-06_10-00.csv",sep=""))
presence_0613 <- read.csv(paste(PATH,"presence_RO_2018-10-06_13-00.csv",sep=""))
presence_0616 <- read.csv(paste(PATH,"presence_RO_2018-10-06_16-00.csv",sep=""))
presence_0619 <- read.csv(paste(PATH,"presence_RO_2018-10-06_19-00.csv",sep=""))
presence_0621 <- read.csv(paste(PATH,"presence_RO_2018-10-06_21-00.csv",sep=""))
presence_0710 <- read.csv(paste(PATH,"presence_RO_2018-10-07_10-00.csv",sep=""))
presence_0713 <- read.csv(paste(PATH,"presence_RO_2018-10-07_13-00.csv",sep=""))
presence_0716 <- read.csv(paste(PATH,"presence_RO_2018-10-07_16-00.csv",sep=""))
presence_0719 <- read.csv(paste(PATH,"presence_RO_2018-10-07_19-00.csv",sep=""))
presence_0721 <- read.csv(paste(PATH,"presence_RO_2018-10-07_21-00.csv",sep=""))

presence_0610$date = "2018-10-06 10h"
presence_0613$date = "2018-10-06 13h"
presence_0616$date = "2018-10-06 16h"
presence_0619$date = "2018-10-06 19h"
presence_0621$date = "2018-10-06 21h"
presence_0710$date = "2018-10-07 10h"
presence_0713$date = "2018-10-07 13h"
presence_0716$date = "2018-10-07 16h"
presence_0719$date = "2018-10-07 19h"
presence_0721$date = "2018-10-07 21h"


presence_0610$crt = 6.10
presence_0613$crt = 6.13
presence_0616$crt = 6.16
presence_0619$crt = 6.19
presence_0621$crt = 6.21
presence_0710$crt = 7.10
presence_0713$crt = 7.13
presence_0716$crt = 7.16
presence_0719$crt = 7.19
presence_0721$crt = 7.21
```

There are (at the end of the vote) totally 10 data files provided with the vote presence, 5 for each day (October 6 and 7), at 10, 13, 16, 19 and 21h.  

The vote is open from 7 to 21 so the data is sampling the vote presence at even intervals.


## Read GeoJSON

We will use a GeoJSON for map representation of Romanian counties.

```{r read_geo_json,warnings=FALSE}
#geoJSON used for rendering, has lower number of points
rgeojson <- readOGR(paste(PATH,"romania.geojson",sep=""))#, layer="OGRGeoJSON")
```

## Read county codes mapping

The county names used in the GeoJSON need to be mapped with the county codes used in the electoral process reports. This additional file is doing this mapping.

```{r read_county_codes_mapping,warnings=FALSE}
county_names_codes <- read.csv(paste(PATH,"county_codes.csv",sep=""))
```

## Read Romania shapefile

```{r read_shapefile,warnings=FALSE}
roshp <- readOGR(paste(PATH,"ro_uat_poligon.shp",sep=""))
```

## Utilities

The function 'matchingNames' is used to do fuzzy matching between Romanian counties names from data files where diferent encoding or writing of these names makes simple matching not  effective.

```{r matching_names}
# Matching names using: https://www.r-bloggers.com/fuzzy-string-matching-a-survival-skill-to-tackle-unstructured-information/
#match using Levenshtein distance between the compared strings 
matchingNames <- function(source1, source2) {
  # To make sure we are dealing with characters
  source1<-as.character(source1)
  source2<-as.character(source2)

  # It creates a matrix with the Standard Levenshtein distance between the name fields of both sources
  dist.name<-adist(source1,source2, partial = FALSE, ignore.case = FALSE)
  # We now take the pairs with the minimum distance
  min.name<-apply(dist.name, 1, min)

  match.s1.s2<-NULL  
  for(i in 1:nrow(dist.name))
  {
    s2.i<-match(min.name[i],dist.name[i,])
    s1.i<-i
    match.s1.s2<-rbind(data.frame(s2.i=s2.i,s1.i=s1.i,
                                  s2name=source2[s2.i], s1name=source1[s1.i], 
                                  adist=min.name[i]),match.s1.s2)
  }

  mapi<-c(1:nrow(dist.name))
  for(j in 1:nrow(dist.name))
  {
    mapi[match.s1.s2$s1.i[j]] = match.s1.s2$s2.i[j]
  }
  return(mapi)
}
```



# **Visualise the data**


## Vote evolution per county


We show first the evolution of the vote (at different hours during the two days), grouped by counties, split on urban and rural areas.


```{r rbubbly,warnings=FALSE}
rbubbly <- function(dataset, x_column, y_column, bubble_column, color_column,
                    slider_column, text_column,
                    x_title=NA, y_title=NA, title=NA,
                    x_is_log=TRUE, y_is_log=FALSE,scale_bubble=1) {

  library(plotly)

  if(x_is_log)
    xtype="log"
  else
    xtype=NA

  if(y_is_log)
    ytype="log"
  else
    ytype=NA

  xmin = min(x_column)
  xmax = max(x_column)
  ymin = min(y_column)
  ymax = max(y_column)

  smin = scale_bubble * sqrt(xmin*ymin)
  smax = scale_bubble * sqrt(xmax*ymax)
  p <-
    plot_ly(data = dataset,
      x = x_column,
      y = y_column,
      size = bubble_column,
      color = color_column,
      frame = slider_column,
      text = text_column,
      hoverinfo = " ~text",
      type = 'scatter',
      mode = 'markers',
      marker = list(opacity = 0.5),
      sizes = c(smin, smax)
    ) %>%
    layout(
      title = title,
      xaxis = list(
        title = x_title,
        type = xtype
      ),
      yaxis = list(
        title = y_title,
        type = ytype
      )
    )
  return (p)
}
```

We will rename the columns of the datasets, after binding the rows (to translate them in English).

```{r smart_bind,warnings=FALSE}
presence = smartbind(presence_0610, presence_0613, presence_0616, presence_0619, presence_0621, presence_0710, presence_0713, presence_0716, presence_0719, presence_0721)

colnames(presence) <- c("County", "Municipality", "City", "SIRUTA", "Circumscription", "Location", "Environment", "Electors", "Votes.Registered", "Votes.not.Registered", "Votes.Mobile", "Votes.Total", "Date.Hour", "Time")
presence = merge(presence,county_names_codes, by.x="County", by.y="Code")

```

Prepare the aggregated data for visualization using Plotly, using a Gapminder-like representation.

```{r rbubbly_prepare,warnings=FALSE}
presence %>% group_by(Time, County, Name, Environment) %>% 
  summarise(Registered.Percent = 100* sum(Votes.Registered)/sum(Electors),
  Not.Registered.Percent = 100 * sum(Votes.not.Registered)/sum(Electors),
  Mobile.Percent = 100 * sum(Votes.Mobile) / sum(Electors),
  Total.Percent = 100 * sum(Votes.Total)/sum(Electors),
  Total.Electors = sum(Electors)) %>% ungroup() -> presence_grouped
```


Each county is represented with a bubble. The size of the bubble is proportional with the total number of electors per county (rural or urban area). On the two axis we show the percent from the total number of electors per county of the registered voters and unregistered voters, respectively.

Data is grouped as well on Rural/Urban electors environment and on sample date/time. 

```{r rbubbly_show,warnings=FALSE}
rbubbly(presence_grouped,
        x_column = presence_grouped$Registered.Percent, y_column = presence_grouped$Total.Percent,
        bubble_column = presence_grouped$Total.Electors, color_column = presence_grouped$Environment,
        slider_column = presence_grouped$Time, text_column = presence_grouped$Name,
        x_title="Percent of registered votes from registered electors", y_title="Percent of total votes from registered electors",                    
        title="Referendum 2018 Romania\nPercent of votes per county, grouped by Rural/Urban environment",
        x_is_log=FALSE, y_is_log=FALSE,scale_bubble=100)
```

We can observe the faster growth of percent of total votes in the rural area for several counties. The counties with largest percent of total votes are both very poor and with a high density of practicant christians.

## Votes evolution per country

Let's show as well the total votes percents per country.


```{r rbubbly_prepare2,warnings=FALSE}
presence %>% group_by(Date.Hour) %>% 
  summarise(Registered.Percent = 100* sum(Votes.Registered)/sum(Electors),
  Not.Registered.Percent = 100 * sum(Votes.not.Registered)/sum(Electors),
  Mobile.Percent = 100 * sum(Votes.Mobile) / sum(Electors),
  Total.Percent = 100 * sum(Votes.Total)/sum(Electors),
  Total.Votes = sum(Votes.Total)) %>% ungroup() -> presence_grouped_all
```



```{r votes_per_country_plotly}

plot_ly(presence_grouped_all, x = ~Date.Hour) %>%
  add_trace(y = ~Total.Percent, name = 'Total %', showlegend=TRUE, type = 'scatter', mode = 'lines+markers') %>%
  add_trace(y = ~Registered.Percent, name = 'Reg. %', showlegend=TRUE, type = 'scatter', mode = 'lines+markers') %>%
  add_trace(y = ~Not.Registered.Percent, name = 'Non Reg. %', showlegend=TRUE, type = 'scatter', mode = 'lines+markers') %>%
  add_trace(y = ~Mobile.Percent, name = 'Mobile %', showlegend=TRUE, type = 'scatter', mode = 'lines+markers') %>%
  layout(title="Percent of the votes from the total registered electors",
         xaxis = list(title = "Date.Time",showticklabels = TRUE, tickangle = 15, tickfont = list(size = 8)),
         yaxis = list(title = "Percent"),
         hovermode = 'compare')  
```

And let's also show the total votes per country, grouped by environment/area (Rural/Urban).

```{r rbubbly_prepare212,warnings=FALSE}
presence %>% group_by(Date.Hour, Environment) %>% 
 summarise(Total.Votes = sum(Votes.Total)) %>% ungroup() -> presence_grouped_all_Environment
```

```{r votes_per_country_plotly212}

plot_ly(presence_grouped_all_Environment, x = ~Date.Hour, color = ~Environment) %>%
  add_trace(y = ~Total.Votes, showlegend=TRUE, type = 'scatter', mode = 'lines+markers') %>%
  layout(title="Total votes per country (Rural(R) and Urban(U) areas)",
         xaxis = list(title = "Date.Time",showticklabels = TRUE, tickangle = 15, tickfont = list(size = 8)),
         yaxis = list(title = "Total"),
         hovermode = 'compare')  
```
We can observe that on the second day of the Referendum, there was a massive mobilization in rural areas (after Oct 7th, 16 pm). This is highly atypical for Romania, where mostly urban dwelers are showing-up for voting in the afternoon of a vote day.   
We cannot speculate on this, but considering that the sudden growth was in that areas that are the stronghold of the Government Party, which activelly support the Referendum, it might be the result of a mobilization of the local party supporters.


## Final votes distribution per county

### Total, non-registered and mobile cast votes

```{r plot_prepare2,warnings=FALSE}
presence %>% filter(Time == max(presence$Time)) %>% group_by(Name) %>% 
  summarise(Registered.Percent = 100* sum(Votes.Registered)/sum(Electors),
  Not.Registered.Percent = 100 * sum(Votes.not.Registered)/sum(Electors),
  Mobile.Percent = 100 * sum(Votes.Mobile) / sum(Electors),
  Total.Percent = 100 * sum(Votes.Total)/sum(Electors),
  Total.Votes = sum(Votes.Total)) %>% ungroup() -> presence_grouped_county
```

```{r votes_per_county_plotly2}

plot_ly(presence_grouped_county, x = ~Name) %>%
  add_trace(y = ~Total.Percent, name = 'Total %', showlegend=TRUE, type = 'bar') %>%
  add_trace(y = ~Registered.Percent, name = 'Reg. %', showlegend=TRUE, type = 'bar') %>%
  add_trace(y = ~Not.Registered.Percent, name = 'Non Reg. %', showlegend=TRUE, type = 'bar') %>%
  add_trace(y = ~Mobile.Percent, name = 'Mobile %', showlegend=TRUE, type = 'bar') %>%
  layout(title="Percent of the votes from the total registered electors",
         xaxis = list(title = "County",showticklabels = TRUE, tickangle = 45, tickfont = list(size = 9)),
         yaxis = list(title = "Percent"),
         hovermode = 'compare')  
```


### Rural and Urban

```{r rbubbly_prepare3,warnings=FALSE}
presence %>% filter(Time == max(presence$Time)) %>% group_by(Name,Environment) %>% 
  summarise(Registered.Percent = 100* sum(Votes.Registered)/sum(Electors),
  Not.Registered.Percent = 100 * sum(Votes.not.Registered)/sum(Electors),
  Mobile.Percent = 100 * sum(Votes.Mobile) / sum(Electors),
  Total.Percent = 100 * sum(Votes.Total)/sum(Electors),
  Total.Votes = sum(Electors)) %>% ungroup() -> presence_grouped_county_Environment
```

```{r votes_per_county_plotly3}

plot_ly(presence_grouped_county_Environment, x = ~Name, color= ~Environment) %>%
  add_trace(y = ~Total.Percent, showlegend=TRUE, type = 'bar') %>%
  layout(title="Percent of the votes from the total registered electors\nRural(R) and Urban(U) areas",
         xaxis = list(title = "County",showticklabels = TRUE, tickangle = 45, tickfont = list(size = 9)),
         yaxis = list(title = "Percent"),
         hovermode = 'compare')  
```

We can see that in 8 counties, in Rural areas, the percent of total votes surpassed the 30% threshold. In the same time, most of the urban areas had under 20% votes. 


Let's visualize the percent of the votes from the total registered electors (total vs. registered) on the same graph, only for rural areas. We also show the names of the counties on the graph and the regression line.


```{r presence_vote_final, fig.width=9, fig.height=9}
presence_grouped_county_Environment %>% 
  ggplot(aes(x=Registered.Percent, y=Total.Percent, color=Environment)) + guides(fill=FALSE) +
  geom_point() + geom_smooth(method=lm) + theme_bw() + 
  scale_x_continuous(limits = c(0, 40)) + scale_y_continuous(limits = c(0,40)) + 
  geom_label_repel(aes(label = Name),box.padding=0.15, point.padding=0.25,segment.color = 'blue') + 
  labs(x="Registered votes from registered electors [%]", y="Total votes from registered electors [%]", 
       title="Percent of votes from registered electors", subtitle="Romanian Referendum 2018 - Rural and Urban areas separated")
```


## Maximum values per polling station

Let's check what are the extreme (maximum) values per polling station. First, let's check the top of the polling stations with the maximum percent of total votes from the number of registered electors.

```{r polling_station_max_votes_percent,fig.width=9, fig.height=9}
presence %>% filter(Date.Hour == max(Date.Hour)) %>% ungroup() -> presence_l
presence_l$Total.Percent = 100 * presence_l$Votes.Total/presence_l$Electors
presence_l$Non.Registered.Percent = 100 * presence_l$Votes.not.Registered/presence_l$Electors

presence_l %>% arrange(desc(presence_l$Total.Percent)) %>% top_n(20,Total.Percent) %>% 
  ggplot(aes(x=Total.Percent, y=Non.Registered.Percent,color=Environment)) + guides(fill=FALSE) +
  geom_point() + geom_smooth(method=lm) + theme_bw() + 
  scale_x_continuous(limits = c(0, 500)) + scale_y_continuous(limits = c(0,500)) + 
  geom_label_repel(size = 3, aes(label = sprintf("%s (%s, %s)",Location, City, Name)),box.padding=0.15, point.padding=0.25,segment.color = 'blue') +
  labs(x="Total votes from registered electors [%]", y="Not registered votes from registered electors [%]", 
       title="Top 20 polling stations by total votes percent from registered electors", subtitle="Romanian Referendum 2018")


```

The maximum values is obtained in one polling station near a famous monastery in `Neamt` county. Most of the high percent of votes polling stations are in Rural areas, with few exception in Urban areas. 


Let's also compar the results for the Remote cast votes.  


```{r polling_station_max_votes_percent1,fig.width=9, fig.height=9}
presence %>% filter(Date.Hour == max(Date.Hour)) %>% ungroup() -> presence_l
presence_l$Mobile.Percent = 100 * presence_l$Votes.Mobile/presence_l$Electors
presence_l$Non.Registered.Percent = 100 * presence_l$Votes.not.Registered/presence_l$Electors

presence_l %>% arrange(desc(presence_l$Mobile.Percent)) %>% top_n(20,Mobile.Percent) %>% 

  ggplot(aes(x=Mobile.Percent, y=Non.Registered.Percent,color=Environment)) + guides(fill=FALSE) +
  geom_point() + geom_smooth(method=lm) + theme_bw() + 
  scale_x_continuous(limits = c(0, 70)) + scale_y_continuous(limits = c(0,70)) + 
  geom_label_repel(size = 3, aes(label = sprintf("%s (%s)",City, Name)),box.padding=0.15, point.padding=0.25,segment.color = 'blue') +
  labs(x="Non-registered votes from registered electors", y="Mobile casted total votes percent from registered electors", 
       title="Top 20 polling stations cities by Mobile casted votes percent from registered electors", subtitle="Romanian Referendum 2018")


```

## Maximum values per city

Second, let's check the cities with the largest percent of total votes from the number of registered electors.


```{r polling_station_max_votes_percent2,fig.width=9, fig.height=9}
presence %>% filter(Time == max(presence$Time)) %>% group_by(City,Name,Environment) %>% 
  summarise(Registered.Percent = 100* sum(Votes.Registered)/sum(Electors),
  Not.Registered.Percent = 100 * sum(Votes.not.Registered)/sum(Electors),
  Mobile.Percent = 100 * sum(Votes.Mobile) / sum(Electors),
  Total.Percent = 100 * sum(Votes.Total)/sum(Electors),
  Total.Votes = sum(Electors)) %>% ungroup() -> presence_l

presence_l %>% arrange(desc(presence_l$Total.Percent))  %>% top_n(20, Total.Percent) %>% 
  ggplot(aes(x=Total.Percent, y=Not.Registered.Percent, color=Environment)) + guides(fill=FALSE) +
  geom_point() + geom_smooth(method=lm) + theme_bw() + 
  scale_x_continuous(limits = c(0, 250)) + scale_y_continuous(limits = c(0,250)) + 
  geom_label_repel(size = 3, aes(label = sprintf("%s (%s)",City, Name)),box.padding=0.15, point.padding=0.25,segment.color = 'blue') +
  labs(x="Total votes from registered electors [%]", y="Not registered votes from registered electors [%]", 
       title="Top 20 cities by total votes percent from registered electors", subtitle="Romanian Referendum 2018")


```


We can observe that the maximum percent of votes are achieved in two kind of cities: turistic resorts (like Geothermal resort `Baile Felix` or mountain resort `Paltinis`) and in remote rural areas. Most of the cities in the top are actually village compounds (`Comuna` in Romanian) from Rural areas, with only few Urban areas.

## Map views for county results

We show here the map of final vote distribution per county.


```{r county_map_prepare,warnings=FALSE}
presence %>% filter(Date.Hour == max(Date.Hour)) %>% 
  group_by(Name) %>%
  summarise(Registered.Percent = 100* sum(Votes.Registered)/sum(Electors),
    Not.Registered.Percent = 100 * sum(Votes.not.Registered)/sum(Electors),
    Mobile.Percent = 100 * sum(Votes.Mobile) / sum(Electors),
    Total.Percent = 100 * sum(Votes.Total)/sum(Electors),
    Total.Votes = sum(Votes.Total)) %>% ungroup() -> presence_county

```

We merge the county codes and names dataset with the presence per country dataset and with the geojson data.

```{r county_map_prepare2,warnings=FALSE}

presence_n = presence_county[matchingNames(rgeojson$name,as.character(presence_county$Name)),]
```

Next, we show the total number of electors per county, the percent of total votes from number of electors, the percent of the non-registered votes from number of electors and the percent of number of mobile cast votes ("urna mobila") from the number of electors per county.


```{r ess}
plot_map <- function(values, main, title, colormap="Blues") {
  
pal <- colorBin(colormap, values, bins = 8, na.color = "Red")
plot(rgeojson, col = pal(values), main = main)
legend(28,48, title = title, legend = c(min(values), max(values)), 
       col = c(pal(min(values)), pal(max(values))),
       fill=c(pal(min(values)), pal(max(values))), cex=0.6)
}

plot_map(presence_n$Total.Votes, "Total votes", "Total votes per county")
plot_map(round(presence_n$Total.Percent,2), "Total votes percent per county", "Total %",  "Blues")
plot_map(round(presence_n$Total.Percent,2), "Total registered votes percent per county", "Reg. %",  "Greens")
plot_map(round(presence_n$Not.Registered.Percent,2), "Total non registered votes percent per county", "Non reg. %", "Oranges")
plot_map(round(presence_n$Mobile.Percent,2), "Total remote cast votes percent per county", "Remote cast %",  "Reds")

```


```{r plot_shp}
plot(roshp)
```

```{r plot_uat_map}
plot_upat_map <- function(values, main, title, colormap="Blues") {
  
pal <- colorBin(colormap, values, bins = 8, na.color = "Red")
plot(roshp, col = pal(values), main = main)
legend(28,48, title = title, legend = c(min(values), max(values)), 
       col = c(pal(min(values)), pal(max(values))),
       fill=c(pal(min(values)), pal(max(values))), cex=0.6)
}

```


# **Updates**

The results are partially, there will be several updates today, for the 16 pm, 19 pm and 21 pm (final) results.  

**Update 1**: Updated for 07-10-2018 16 pm.  

**Update 2**: Updated for 07-10-2018 19 pm.  

**Update 3**: Updated for 07-10-2018 21 pm.  

# **Conclusion**

The Referendum was not validated, due to total vote attendance under the 30% threshold necessary for validation. Although at the end of the 2nd day a sudden mobilization in around 8 counties brought the rural areas in these counties over 30%, it was not enough, with most urban areas with a total votes percent under 20%.  At the end of the 2nd day of the Referendum, total number of rural areas votes exceeded the total number of urban areas votes.  


# **References**

[1] The World Factbook - Romania, https://www.cia.gov/library/publications/the-world-factbook/geos/ro.html  
[2] Central Electoral Bureau of Romania, Data Source for the vote presence (in RO lang), http://prezenta.bec.ro/  
[3] Romanian Geospatial Data, Data Source for GeoJSON with Romanian counties (in RO lang), http://www.geo-spatial.org/download/romania-seturi-vectoriale

