---
title: "R Elo EDA"
author: "Gabriel Preda"
date: "Created: 2018-12-28; Last updated: `r Sys.Date()`"
output:
  html_document:
    number_sections: false
    toc: true
    fig_width: 8
    fig_height: 6
    theme: cosmo
    highlight: tango
    code_folding: hide
---

#**Introduction** 


This Kernel will take you through the process of analyzing the data to understand the predictive values of various features and the possible correlation between different features.


#**Prepare the data analysis**

##Load packages  

Let's load the packages.

```{r,message=FALSE,warning=FALSE}
library(caret)
library(corrplot)
library(readr)
library(knitr)
library(kableExtra)
library(formattable)
library(dplyr)
library(ggplot2)
library(gridExtra)
library(grid)
library(plotly)
options(knitr.table.format = "html") 
```  


##Load the data

Let's load the data.


```{r,input_data, message=FALSE,warning=FALSE}
PATH="../input/elo-merchant-category-recommendation/"
#PATH="../input"
train_df=read_csv(paste0(PATH,'train.csv'))
test_df=read_csv(paste0(PATH,'test.csv'))
historical_trans_df=read_csv(paste0(PATH,'historical_transactions.csv'))
new_merchant_trans_df=read_csv(paste0(PATH,'new_merchant_transactions.csv'))
merchant_df=read_csv(paste0(PATH,'merchants.csv'))
```  

#**Summary of the data**{.tabset .tabset-fade .tabset-pills}


We loaded 5 data files from the dataset, as following:

* train.csv (`r nrow(train_df)` rows, `r length(names(train_df))` columns)  
* test.csv (`r nrow(test_df)` rows, `r length(names(test_df))` columns)  
* historical_transactions.csv (`r nrow(historical_trans_df)` rows, `r length(names(historical_trans_df))` columns)  
* new_merchant_transactions.csv (`r nrow(new_merchant_trans_df)` rows, `r length(names(new_merchant_trans_df))` columns)
* merchants.csv (`r nrow(merchant_df)` rows, `r length(names(merchant_df))` columns)


Let's see the first few rows of each data file and also glimpse these data files.


##Train  

```{r glimpse_train}
knitr::kable(head(train_df,10),caption="Train (first 10 rows)")
glimpse(train_df)
```

##Test

```{r glimpse_test}
knitr::kable(head(test_df,10),caption="Test (first 10 rows)")
glimpse(test_df)
```


##Historical transactions

```{r glimpse_historical_transactions}
knitr::kable(head(historical_trans_df,10),caption="Historical transactions (first 10 rows)")
glimpse(historical_trans_df)
```

##New merchant transactions 

```{r glimpse_new_merchant_transactions}
knitr::kable(head(new_merchant_trans_df,10),caption="New merchant transactions (first 10 rows)")
glimpse(new_merchant_trans_df)
```


##Merchants


```{r glimpse_merchants}
knitr::kable(head(merchant_df,10),caption="Merchants (first 10 rows)")
glimpse(merchant_df)
```


#**Data exploration**

Let's check for missing data in dataframes.   

##Check for missing data

We will define a function to plot the percent of missing values per each column of the dataframe.

```{r function_missing_data}
missing_data <- function(data) {
  apply(data, 2, function(col)sum(is.na(col))/length(col))
}

print("Train data"); missing_data(train_df)
print("Test data"); missing_data(test_df)
print("Historical transaction data"); missing_data(historical_trans_df)
print("New merchant transaction data"); missing_data(new_merchant_trans_df)
print("Merchants data"); missing_data(merchant_df)

```


##Train and test data  

Let's check the distribution of train and test features.

Both have the same features:

* card_id;  
* feature1, feature2, feature3;  
* first_active_month;  

Train has also the target value, called target.


```{r plot_bar}
plot_bar <- function(data, var, name='Number', color='blues',
                     title='Number', xtitle='feature', ytitle='Number of records',
                     width=400, height=400) {
  var <- enquo(var)
  
  data %>% group_by(!!var) %>% summarise(nr = length(card_id)) %>% ungroup() -> data_group
  plot_ly(data_group, x = ~var,width = width, height = height) %>%
    add_trace(y = ~nr, name=name, marker=list(color=color), showlegend=TRUE, type = 'bar') %>%
  layout(title=title,
         xaxis = list(title = xtitle, showticklabels = TRUE, tickangle = 45, tickfont = list(size = 8)),
         yaxis = list(title = ytitle),
         hovermode = 'compare')  
  }
```


```{r feature_1}
plot_bar(train_df, feature_1,xtitle='Feature 1', color='blue', title='Train data')
plot_bar(test_df, feature_1,xtitle='Feature 1', color='lightblue', title='Test data')

ggplot(train_df, aes(x=target)) +  theme_bw() +
  geom_density(aes(group=feature_1, color=feature_1))
```

```{r feature_2}
plot_bar(train_df, feature_2,xtitle='Feature 2', color='blue', title='Train data')
plot_bar(test_df, feature_2,xtitle='Feature 2', color='lightblue', title='Test data')

ggplot(train_df, aes(x=target)) +  theme_bw() +
  geom_density(aes(group=feature_2, color=feature_2))
```

```{r feature_3}
plot_bar(train_df, feature_3,xtitle='Feature 3', color='blue', title='Train data')
plot_bar(test_df, feature_3,xtitle='Feature 3', color='lightblue', title='Test data')

ggplot(train_df, aes(x=target)) + theme_bw() +
  geom_density(aes(group=feature_3, color=feature_3))
```

```{r first_active_month}
plot_bar(train_df, first_active_month,xtitle='First active month', color='blue', title='Train data')
plot_bar(test_df, first_active_month,xtitle='First active month', color='lightblue', title='Test data')
```



##Historical transaction data

Let's check the distribution of historical transaction data features.

historical_trans_df is linked with train_df and test_df by the card_id key.

Let's plot category_1, category_2, category_3 features distribution.


```{r category_1}
plot_bar(historical_trans_df, category_1,xtitle='Category 1', color='magenta', title='Historical transaction data')
```

```{r category_2}
plot_bar(historical_trans_df, category_2,xtitle='Category 2', color='red', title='Historical transaction data')
```

```{r category_3}
plot_bar(historical_trans_df, category_3,xtitle='Category 3', color='yellow', title='Historical transaction data')
```


Let's see city_id, merchant_category_id, state_id, subsector_id.


```{r city_id}
plot_bar(historical_trans_df, city_id,xtitle='City ID', color='lightblue', title='Historical transaction data')
```


```{r merchant_category_id}
plot_bar(historical_trans_df, merchant_category_id,xtitle='Merchant category ID', color='darkblue', title='Historical transaction data')
```

```{r state_id}
plot_bar(historical_trans_df, state_id,xtitle='State ID', color='tomato', title='Historical transaction data')
```

```{r subsector_id}
plot_bar(historical_trans_df, subsector_id,xtitle='Subsector ID', color='darkgreen', title='Historical transaction data')
```


Let's show the purchase amount grouped by purchase time types.

Before this, let's extract the date.

historical_trans_df['purchase_date'] = pd.to_datetime(historical_trans_df['purchase_date'])
historical_trans_df['month'] = historical_trans_df['purchase_date'].dt.month
historical_trans_df['dayofweek'] = historical_trans_df['purchase_date'].dt.dayofweek
historical_trans_df['weekofyear'] = historical_trans_df['purchase_date'].dt.weekofyear

```{r extract_date_time}
dtime <- strptime(str1, "%Y-%m-%dT%H:%M:%SZ")

```

##New merchant transaction data

Let's check the distribution of historical transaction data features.

historical_trans_df is linked with train_df and test_df by the card_id key.

Let's plot category_1, category_2, category_3 features distribution.


```{r nm_category_1}
plot_bar(new_merchant_trans_df, category_1,xtitle='Category 1', color='blue', title='New merchant transaction data')
```

```{r nm_category_2}
plot_bar(new_merchant_trans_df, category_2,xtitle='Category 2', color='lightblue', title='New merchant transaction data')
```

```{r nm_category_3}
plot_bar(new_merchant_trans_df, category_3,xtitle='Category 3', color='blue', title='New merchant transaction data')
```


Let's see city_id, merchant_category_id, state_id, subsector_id.


```{r nm_city_id}
plot_bar(new_merchant_trans_df, city_id,xtitle='City ID', color='red', title='New merchant transaction data')
```


```{r nm_merchant_category_id}
plot_bar(new_merchant_trans_df, merchant_category_id,xtitle='Merchant category ID', color='green', title='New merchant transaction data')
```

```{r nm_state_id}
plot_bar(new_merchant_trans_df, state_id,xtitle='State ID', color='darkgrey', title='New merchant transaction data')
```

```{r nm_subsector_id}
plot_bar(historical_trans_df, subsector_id,xtitle='Subsector ID', color='magenta', title='New merchant transaction data')
```


#**Final note**

This Kernel is still under construction.